Gravitational wave sources in our Galactic backyard: LISA predictions for BH and NS binaries

### 2021-06-17 paper/tex/sections/results.tex
move plots around
results proofread

\subsubsection{Eccentricity uncertainty}\label{sec:ecc_unc}
One of the main advantage of space-based gravitational wave detectors such as LISA is that systems may still have significant eccentricity during the LISA mission. Whilst also being useful a useful individual quantity for learning more about sources, the uncertainty on the eccentricity also affects the uncertainty on the chirp mass and so it is important to quantify this uncertainty.

@@

In some cases, sources may only have one individually detectable harmonics, or even none. In these cases one can only put a lower or upper bound on the eccentricity. For sources with only one detectable harmonic, the source is either a circular source or a source with a small enough eccentricity that most of the GW power is still concentrated in the $n = 2$. Therefore, we can use this information to place an upper bound on the eccentricity. Conversely, sources with no detectable harmonics must be eccentric such that the GW power has spread over so many harmonics that no single harmonic has a significantly stronger signal. Therefore, we can use this information to place a lower bound on the eccentricity.

This method provides a pessimistic estimate of the eccentricity of sources since we do not consider the benefits of any sort of matched-filter analysis or similar methods. We discuss how the pessimism of this method affects our results and some possible alternative in Section~\ref{sec:caveats}.

\subsubsection{Chirp mass uncertainty}
The chirp mass uncertainty can be calculated using the uncertainty on the orbital frequency, the time derivative of the orbital frequency and the eccentricity. This is because the time derivative of the orbital frequency can be written as
\begin{equation}
    \dot{f}_{\rm orb} = \frac{48}{5 \pi} \frac{(G \mathcal{M}_c)^{5/3}}{c^3} (2 \pi f_{\rm orb})^{11/3} F(e),
\end{equation}
where $f_{\rm orb}$ is the orbital frequency, $\mathcal{M}_{c}$ is the chirp mass (defined in Eq.~\ref{eq:chirp_mass}) and $e$ is the eccentricity and
\begin{equation}
    F(e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}},
\end{equation}
is the enhancement factor of gravitational wave emission for an eccentric binary over an otherwise identical circular binary \citep[][Eq.~17]{Peters+1963}.

We can invert this and instead write that the chirp mass is
\begin{equation}
    \mathcal{M}_c = \frac{c^3}{G} \left( \frac{5 \pi}{48} \frac{\dot{f}_{\rm orb}}{F(e)} \right)^{3/5} \frac{1}{(2 \pi f_{\rm orb})^{11/5}},
\end{equation}
and therefore that the chirp mass uncertainty is
\begin{equation}\label{eq:chirp_mass_uncertainty}
    \qty( \frac{\Delta \mathcal{M}_c}{\mathcal{M}_c} )^2 = \left( \frac{11}{5} \frac{\Delta f_{\rm orb}}{f_{\rm orb}} \right)^2 + \left( \frac{3}{5} \frac{\Delta \dot{f}_{\rm orb}}{\dot{f}_{\rm orb}} \right)^2 + \left( \frac{3}{5} \frac{\Delta F(e)}{F(e)} \right)^2.
\end{equation}

We calculate the frequency uncertainties using \citet{Takahashi+2002}, such that
\begin{align}\label{eq:f_orb_unc}
    \frac{\Delta f_{\rm orb}}{f_{\rm orb}} &= 4 \sqrt{3} \cdot \frac{1}{\rho} \frac{1}{T_{\rm obs}} \frac{1}{f_{\rm orb}}, \\
    \frac{\Delta \dot{f}_{\rm orb}}{\dot{f}_{\rm orb}} &= 6 \sqrt{5} \cdot \frac{1}{\rho} \left(\frac{1}{T_{\rm obs}} \right)^2 \frac{1}{\dot{f}_{\rm orb}},
\end{align}
where $\rho$ is the signal-to-noise ratio and $T_{\rm obs}$ is the LISA mission length. Then we calculate the eccentricity certainty as discussed in Sec.~\ref{sec:ecc_unc} and propagate it so that
\begin{equation}
    \frac{\Delta F(e)}{F(e)} = \Delta e \cdot \frac{(1256 + 1608 e^2 + 111 e^4) e}{96 + 196 e^2 - 255 e^4 - 37 e^6}.
\end{equation}

We use Eq.~\ref{eq:chirp_mass_uncertainty} to calculate the chirp mass uncertainty for each DCO in our sample and plot it in the lower centre panel of Fig.~\ref{fig:fiducial_pdf_distributions}. We show the distribution as a CDF that is normalised to the expected number of detections to better indicate the absolute number of detections that will have well determined chirp masses. We find that approximately 10 BHBHs, 10 BHNSs and 5 NSNSs have measurable chirp masses (as indicated by the shaded region). We emphasise that this is a pessimistic estimate due to the pessimistic nature of how we determine the eccentricity uncertainty.

>>>>

\subsubsection{Eccentricity uncertainty}\label{sec:ecc_unc}
One of the main advantage of space-based gravitational wave detectors such as LISA is that systems may still have significant eccentricity during the LISA mission. Whilst also being useful as an individual quantity for learning more about sources, the uncertainty on the eccentricity also affects the uncertainty on the chirp mass and so it is important to quantify this uncertainty.

@@

In some cases, sources may only have one individually detectable harmonic, or even none. In these cases one can only put a lower or upper bound on the eccentricity. For sources with only one detectable harmonic, the source is either a circular source or a source with a small enough eccentricity that most of the GW power is still concentrated in the $n = 2$ harmonic. Therefore, we can use this information to place an upper bound on the eccentricity. Conversely, sources with no detectable harmonics must be eccentric enough such that the GW power has spread over so many harmonics that no single harmonic has a significantly stronger signal. Therefore, we can use this information to place a lower bound on the eccentricity.

This method provides a pessimistic estimate of the eccentricity of sources since we do not consider the benefits of any sort of matched-filter analysis or similar methods.

\subsubsection{Chirp mass uncertainty}
The chirp mass uncertainty can be calculated using the uncertainty on the orbital frequency, the time derivative of the orbital frequency and the eccentricity. This is because the time derivative of the orbital frequency can be written as
\begin{equation}\label{eq:f_n_dot}
    \dot{f}_{n} = \frac{48 n}{5 \pi} \frac{(G \mathcal{M}_c)^{5/3}}{c^3} (2 \pi f_{\rm orb})^{11/3} F(e),
\end{equation}
where $f_{\rm orb}$ is the orbital frequency, $\mathcal{M}_{c}$ is the chirp mass (defined in Eq.~\ref{eq:chirp_mass}) and $e$ is the eccentricity and
\begin{equation}
    F(e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}},
\end{equation}
is the enhancement factor of gravitational wave emission for an eccentric binary over an otherwise identical circular binary \citep[][Eq.~17]{Peters+1963}.

We can invert this and instead write that the chirp mass is
\begin{equation}
    \mathcal{M}_c = \frac{c^3}{G} \left( \frac{5 \pi}{48 n} \frac{\dot{f}_{n}}{F(e)} \right)^{3/5} \frac{1}{(2 \pi f_{\rm orb})^{11/5}},
\end{equation}
and therefore that the chirp mass uncertainty is
\begin{equation}\label{eq:chirp_mass_uncertainty}
    \qty( \frac{\Delta \mathcal{M}_c}{\mathcal{M}_c} )^2 = \left( \frac{11}{5} \frac{\Delta f_{\rm orb}}{f_{\rm orb}} \right)^2 + \left( \frac{3}{5} \frac{\Delta \dot{f}_{\rm dom}}{\dot{f}_{\rm dom}} \right)^2 + \left( \frac{3}{5} \frac{\Delta F(e)}{F(e)} \right)^2,
\end{equation}
where $f_{\rm dom}$ is the harmonic frequency with the strongest SNR as this will provide the best measurement.

We calculate the frequency uncertainties using \citet{Takahashi+2002}, such that
\begin{align}\label{eq:f_orb_unc}
    \frac{\Delta f_{\rm orb}}{f_{\rm orb}} &= 4 \sqrt{3} \cdot \frac{1}{\rho} \frac{1}{T_{\rm obs}} \frac{1}{f_{\rm orb}}, \\
    \frac{\Delta \dot{f}_{\rm dom}}{\dot{f}_{\rm dom}} &= 6 \sqrt{5} \cdot \frac{1}{\rho} \left(\frac{1}{T_{\rm obs}} \right)^2 \frac{1}{\dot{f}_{\rm dom}},
\end{align}
where $\rho$ is the signal-to-noise ratio and $T_{\rm obs}$ is the LISA mission length. Then we calculate the eccentricity certainty, $\Delta e$, as discussed in Sec.~\ref{sec:ecc_unc} and propagate it so that
\begin{equation}
    \frac{\Delta F(e)}{F(e)} = \Delta e \cdot \frac{(1256 + 1608 e^2 + 111 e^4) e}{96 + 196 e^2 - 255 e^4 - 37 e^6}.
\end{equation}

We use Eq.~\ref{eq:chirp_mass_uncertainty} to calculate the chirp mass uncertainty for each DCO in our sample and plot it in the lower centre panel of Fig.~\ref{fig:fiducial_pdf_distributions}. We show the distribution as a CDF that is normalised to the expected number of detections to better indicate the absolute number of detections that will have well determined chirp masses. We find that approximately 10 BHBHs, 10 BHNSs and 5 NSNSs have measurable chirp masses (as indicated by the shaded region). This uncertainty is generally dominated by the uncertainty on the time derivative of the frequency since most of the binaries are too early in their inspiral for LISA to measure a strong chirp.
